---
title: "Analysis: Cross-Country Portrayal of Outer Space in Sci-Fi"
author: "Donald Moratz"
date: "2026-05-26"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    fig_width: 10
    fig_height: 7
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
    echo    = TRUE,
    message = FALSE,
    warning = FALSE
)

# Wipe the environment so every knit / run starts from a known state.
rm(list = ls())

# Load shared setup (packages, paths, helpers) plus analysis-specific
# helpers.  Both files live alongside this .Rmd in scripts/.
source("source.R")
source("source_analysis.R")

```


# Purpose

Descriptive, hypothesis-generating analysis of how sci-fi novels from
six countries (China, Finland, France, Japan, Russia, US) portray outer
space.  The pipeline rolls Claude's per-chapter codings up to the book
and country level using the rules locked in for Task 3
(PROJECT_TASKS.md §5):

* **Chapter -> Book**: majoritarian (mode) coding with the "any non-
  Other / Unsure chapter wins" rule.  Computed in parallel: mode,
  median, mean, peak, end.  Tie-breaking via median-of-tied-positions
  for ordinal questions, lower-index canonical option for nominal
  questions.
* **Book -> Country**: weighted aggregation that lets collection-heavy
  countries fairly contribute.  Each story in an N-story collection
  contributes 1/N of a book's weight by default; flip
  `collection_weight_mode` below to change.
* **No inferential testing** -- this is explicitly hypothesis-
  generating analysis for a Nature submission.  Distributions, modes,
  divergences, and trajectories are shown directly.

---

# Analysis configuration

The two main analytical levers sit here so collaborators can change
them without editing `source.R`.

```{r analysis-config}

# Which book -> country aggregation mode to use.  Locked-in default is
# "weighted"; flip to "equal" to treat every story as a full book, or
# to "merged" to pool collection chapters back into single books.
collection_weight_mode <- "weighted"

# Where trajectory + heatmap PNGs get written for downstream use.
FIGURE_DIR <- file.path(PROCESSED_DIR, "figures")
if (!dir.exists(FIGURE_DIR)) dir.create(FIGURE_DIR, recursive = TRUE)

cat("collection_weight_mode:", collection_weight_mode, "\n")
cat("FIGURE_DIR:            ", FIGURE_DIR, "\n")

```


# Inputs

## Load questions, Claude long frame, and collection metadata

```{r load-inputs}

questions <- jsonlite::read_json(QUESTIONS_JSON, simplifyVector = FALSE)
claude_long <- readRDS(file.path(PROCESSED_DIR, "space_chapters_long.rds"))
collection_meta <- load_collection_metadata()

cat("Questions:        ", length(questions), "\n")
cat("Claude long rows: ", nrow(claude_long), "\n")
cat("Distinct books:   ", dplyr::n_distinct(claude_long$book), "\n")
cat("Distinct countries:", dplyr::n_distinct(claude_long$country), "\n")

cat("\nCollection metadata summary:\n")
collection_meta %>%
    dplyr::count(classification, is_collection) %>%
    knitr::kable()

```


## Attach ordinal positions

```{r add-ordinal}

claude_long <- add_ordinal_position(claude_long, questions)

# Quick sanity check: every committed ordinal answer should have an
# ordinal_position; every Other / Unsure should have NA.
claude_long %>%
    dplyr::filter(question_number %in% ordinal_question_numbers) %>%
    dplyr::count(has_pos = !is.na(ordinal_position),
                 is_other = answer == "Other / Unsure") %>%
    knitr::kable(caption = "Ordinal-position attachment sanity check (ordinal Qs only)")

```


## Apply collection split

```{r split-collections}

claude_long_split <- split_collections(claude_long, collection_meta,
                                       mode = collection_weight_mode)

cat("Rows after split:        ", nrow(claude_long_split), "\n")
cat("Distinct books after split:",
    dplyr::n_distinct(claude_long_split$book), "\n")

claude_long_split %>%
    dplyr::filter(!is.na(parent_book)) %>%
    dplyr::distinct(country, parent_book, book) %>%
    dplyr::count(country, parent_book, name = "n_stories") %>%
    knitr::kable(caption = "Stories per split collection")

```


# Chapter -> Book aggregation

We produce two book-level frames:

* **`book_level_split`** -- 106 rows.  Each chapter of a collection is
  its own book row (with `parent_book` set).  This is what the country-
  level aggregation uses so the collection-weighting parameter works.
* **`book_level`** (headline) -- 46 rows, one per parent book.  Built
  by collapsing collection rows in `book_level_split` via
  `collapse_collections_to_books()`, which applies the mode-of-per-
  story-modes rule with the same tie-breaking we use at the chapter ->
  book layer.

```{r aggregate-books}

book_level_split <- aggregate_chapter_to_book(claude_long_split, questions)

book_level <- collapse_collections_to_books(book_level_split, questions)

cat("Split book-level rows:    ", nrow(book_level_split), "\n")
cat("Distinct books (split):   ", dplyr::n_distinct(book_level_split$book), "\n")
cat("Collapsed book-level rows:", nrow(book_level), "\n")
cat("Distinct books (headline):", dplyr::n_distinct(book_level$book), "\n")

book_level %>%
    dplyr::count(tie_broken, name = "n") %>%
    knitr::kable(caption = "Tie-broken book codings in the headline frame (low count is good)")

book_level %>%
    dplyr::count(all_other_unsure, name = "n") %>%
    knitr::kable(caption = "Headline books that fell back to Other / Unsure (all chapters / stories were Other / Unsure)")

```


## Where do mode and peak diverge?

Cases where `mode_answer != peak_answer` for ordinal questions are
exactly the books where Donald's "single peak chapter pulls memory"
intuition could matter most.

```{r mode-peak-divergence}

book_level %>%
    dplyr::filter(scale == "ordinal",
                  !is.na(mode_answer), !is.na(peak_answer),
                  mode_answer != peak_answer) %>%
    dplyr::transmute(
        country, book,
        q_label = q_label_short(question_number),
        mode_short = get_answer_short(question_number, mode_answer),
        peak_short = get_answer_short(question_number, peak_answer),
        peak_minus_mode = peak_position - mode_position
    ) %>%
    dplyr::arrange(dplyr::desc(peak_minus_mode)) %>%
    head(40) %>%
    knitr::kable(caption = "Top 40 mode-vs-peak divergences (ordinal Qs)")

```


# Book -> Country aggregation

The country roll-up uses the **split** book-level frame so the
`collection_weight_mode` parameter can apply its 1/N down-weighting per
collection story.  The headline 46-row collapsed frame is for plotting
only.

```{r aggregate-countries}

country_agg <- aggregate_book_to_country(
    book_level_split,
    collection_meta,
    weight_mode = collection_weight_mode
)

country_distributions <- country_agg$distributions
country_aggregates    <- country_agg$aggregates

cat("Country distribution rows:", nrow(country_distributions), "\n")
cat("Country aggregate rows:   ", nrow(country_aggregates), "\n")

country_aggregates %>%
    dplyr::filter(scale == "ordinal") %>%
    dplyr::select(country, question_number, n_books, total_weight,
                  mean_position, median_position, mean_peak_position) %>%
    dplyr::mutate(dplyr::across(c(mean_position, median_position,
                                   mean_peak_position),
                                ~ round(.x, 2))) %>%
    head(30) %>%
    knitr::kable(caption = "Country-level ordinal aggregates (first 30 rows)")

```


# Visualizations

## Reference chart (color key for all subsequent plots)

All ordinal questions (Q01-Q08, Q10, Q12) use the same 5-color
gradient, with position 1 representing "safe / familiar / orderly"
and position 5 representing "extreme / alien / chaotic".  Q02
(inhabitable) and Q07 (commonness) were inverted vs the original
questions.json order, and Q08 (polities) was reordered, to make the
gradient consistent across questions.  Q09 and Q11 are nominal; each
category has its own color shown below.  "Other / Unsure" is grey
everywhere.

```{r reference-chart, fig.width = 13, fig.height = 9}

p_ref <- plot_legend_reference(questions)
print(p_ref)

ggplot2::ggsave(file.path(FIGURE_DIR, "reference_chart.png"),
                p_ref, width = 13, height = 9, dpi = 120)

```


## Book fingerprint heatmaps (mode + peak)

Headline view: 46 rows, one per parent book.  Collections are
collapsed via mode-of-per-story-modes.  The full per-story (106-row)
breakouts live in the Supplementary Analysis section near the end of
this document.

```{r book-fingerprint, fig.width = 11, fig.height = 12}

p_fp_mode <- plot_book_fingerprint(book_level, questions,
                                   use_field = "mode_answer")
print(p_fp_mode)

ggplot2::ggsave(file.path(FIGURE_DIR, "book_fingerprint_mode.png"),
                p_fp_mode, width = 11, height = 12, dpi = 120)

```

```{r book-fingerprint-peak, fig.width = 11, fig.height = 12}

p_fp_peak <- plot_book_fingerprint(book_level, questions,
                                   use_field = "peak_answer")
print(p_fp_peak)

ggplot2::ggsave(file.path(FIGURE_DIR, "book_fingerprint_peak.png"),
                p_fp_peak, width = 11, height = 12, dpi = 120)

```


## Country distribution stacked bars (mode + peak)

```{r country-distribution-mode, fig.width = 12, fig.height = 11}

p_dist_mode <- plot_country_distribution(country_distributions, questions,
                                         use_field = "mode")
print(p_dist_mode)

ggplot2::ggsave(file.path(FIGURE_DIR, "country_distribution_mode.png"),
                p_dist_mode, width = 12, height = 11, dpi = 120)

```

```{r country-distribution-peak, fig.width = 12, fig.height = 11}

p_dist_peak <- plot_country_distribution(country_distributions, questions,
                                         use_field = "peak")
print(p_dist_peak)

ggplot2::ggsave(file.path(FIGURE_DIR, "country_distribution_peak.png"),
                p_dist_peak, width = 12, height = 11, dpi = 120)

```


## Country divergence heatmaps (mode + peak)

```{r country-divergence-mode, fig.width = 10, fig.height = 5}

p_div_mode <- plot_country_divergence(country_aggregates, questions,
                                      use_field = "mean_position")
print(p_div_mode)

ggplot2::ggsave(file.path(FIGURE_DIR, "country_divergence_mode.png"),
                p_div_mode, width = 10, height = 5, dpi = 120)

```

```{r country-divergence-peak, fig.width = 10, fig.height = 5}

p_div_peak <- plot_country_divergence(country_aggregates, questions,
                                      use_field = "mean_peak_position")
print(p_div_peak)

ggplot2::ggsave(file.path(FIGURE_DIR, "country_divergence_peak.png"),
                p_div_peak, width = 10, height = 5, dpi = 120)

```


# Trajectories

## Per-book trajectories (46 figures)

For each book in the corpus, a 12-panel figure showing how its
chapter-level codings evolve over the course of the narrative.
Collection rows are merged back to parent books for the trajectory
view -- a collection's trajectory is its full sequence of stories.

```{r trajectory-books, fig.width = 11, fig.height = 9, results = "asis"}

# For trajectories we want the ORIGINAL claude_long (not split), so each
# collection shows as one book with its full chapter sequence.
trajectory_books <- claude_long %>%
    dplyr::distinct(country, book) %>%
    dplyr::arrange(country, book)

for (i in seq_len(nrow(trajectory_books))) {

    b <- trajectory_books$book[i]
    c <- trajectory_books$country[i]

    cat("\n\n### ", c, " - ", b, "\n\n", sep = "")

    p <- plot_trajectory_per_book(claude_long, b, questions)
    print(p)

    safe_name <- gsub("[^A-Za-z0-9]+", "_", b)
    ggplot2::ggsave(
        file.path(FIGURE_DIR, paste0("trajectory_book_", safe_name, ".png")),
        p, width = 11, height = 9, dpi = 100
    )
}

```


## Per-question trajectories (12 figures)

For each question, one figure with a small panel per book showing how
the answer evolves.  Useful for spotting cross-book arc patterns
(e.g., are contestation trajectories systematically escalating in war-
focused books?).

```{r trajectory-questions, fig.width = 14, fig.height = 16, results = "asis"}

for (q in questions) {

    qn <- q$number

    cat("\n\n### ", q_label_short(qn), ": ", q$criteria_short, "\n\n", sep = "")

    p <- plot_trajectory_per_question(claude_long, qn, questions)
    print(p)

    ggplot2::ggsave(
        file.path(FIGURE_DIR,
                  paste0("trajectory_", q_label_short(qn), ".png")),
        p, width = 14, height = 16, dpi = 100
    )
}

```


# Supplementary Analysis

The headline book fingerprint above collapses each collection into a
single row via mode-of-per-story-modes.  In this section we present
each collection's per-story breakout as a dedicated mini-fingerprint,
so within-collection variation is visible for the four collections /
anthologies in the corpus.

```{r supplementary-collections, fig.width = 11, fig.height = 7, results = "asis"}

collection_parents <- collection_meta %>%
    dplyr::filter(is_collection) %>%
    dplyr::arrange(country, book) %>%
    dplyr::pull(book)

for (parent in collection_parents) {

    sub <- book_level_split %>%
        dplyr::filter(parent_book == parent)

    cat("\n\n## ", parent, " (", dplyr::n_distinct(sub$book),
        " stories)\n\n", sep = "")

    p <- plot_book_fingerprint(sub, questions, use_field = "mode_answer")
    print(p)

    safe_name <- gsub("[^A-Za-z0-9]+", "_", parent)
    ggplot2::ggsave(
        file.path(FIGURE_DIR,
                  paste0("supplementary_fingerprint_", safe_name, ".png")),
        p, width = 11, height = 7, dpi = 120
    )
}

```


# Save outputs

```{r save-outputs}

saveRDS(book_level,
        file.path(PROCESSED_DIR, "book_level.rds"))
saveRDS(book_level_split,
        file.path(PROCESSED_DIR, "book_level_split.rds"))
saveRDS(country_distributions,
        file.path(PROCESSED_DIR, "country_distributions.rds"))
saveRDS(country_aggregates,
        file.path(PROCESSED_DIR, "country_aggregates.rds"))

readr::write_csv(book_level,
                 file.path(PROCESSED_DIR, "book_level.csv"))
readr::write_csv(book_level_split,
                 file.path(PROCESSED_DIR, "book_level_split.csv"))
readr::write_csv(country_distributions,
                 file.path(PROCESSED_DIR, "country_distributions.csv"))
readr::write_csv(country_aggregates,
                 file.path(PROCESSED_DIR, "country_aggregates.csv"))

cat("Wrote analysis outputs to", PROCESSED_DIR, "\n")
cat("Wrote", length(list.files(FIGURE_DIR)),
    "figure files to", FIGURE_DIR, "\n")

```


## Generate analysis_report.md

```{r write-report}

report_lines <- c(
    "# Analysis Report",
    "",
    paste0("_Generated: ", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "_"),
    "",
    "## Configuration",
    "",
    paste0("- `collection_weight_mode`: ", collection_weight_mode),
    paste0("- Books in corpus (pre-split): ",
           dplyr::n_distinct(claude_long$book)),
    paste0("- Books in corpus (post-split): ",
           dplyr::n_distinct(claude_long_split$book)),
    paste0("- Countries: ",
           paste(sort(unique(claude_long$country)), collapse = ", ")),
    "",
    "## Book-level coding statistics",
    "",
    paste0("- Total book-level (book x question) rows: ",
           nrow(book_level)),
    paste0("- Rows requiring tie-breaking: ",
           sum(book_level$tie_broken, na.rm = TRUE)),
    paste0("- Books coded Other / Unsure (all chapters were Other / Unsure): ",
           sum(book_level$all_other_unsure, na.rm = TRUE)),
    paste0("- Ordinal books where peak != mode: ",
           sum(book_level$scale == "ordinal" &
                   !is.na(book_level$mode_answer) &
                   !is.na(book_level$peak_answer) &
                   book_level$mode_answer != book_level$peak_answer,
               na.rm = TRUE)),
    "",
    "## Outputs",
    "",
    "- `data/processed/book_level.{rds,csv}`",
    "- `data/processed/country_distributions.{rds,csv}`",
    "- `data/processed/country_aggregates.{rds,csv}`",
    paste0("- `data/processed/figures/` (",
           length(list.files(FIGURE_DIR)), " files)"),
    "",
    "## Carry-forward IRR caveats from Task 2",
    "",
    "- Mean per-question weighted kappa across the 5 IRR books: 0.857 (ordinal questions).",
    "- Mean per-question alpha across the 5 IRR books: 0.703.",
    "- Where both Claude and the human committed to ordinal answers, agreement is strong; pooled nominal agreement is weaker (pooled alpha = 0.389 with all cells, 0.636 excluding one-sided Other / Unsure).",
    "- Q9 (genre) and Q11 (nearest-neighbor metaphor) have weak chapter-level reliability (kappa ~ 0.03-0.06); present their distributions with this caveat.",
    "- Planet of the Apes had heavy human hedging (68% Other / Unsure); the Claude-only analysis here is unaffected but cross-book comparisons that hinge on Apes should be noted."
)

writeLines(
    report_lines,
    file.path(PROCESSED_DIR, "analysis_report.md")
)

cat("Wrote analysis_report.md to", PROCESSED_DIR, "\n")

```


# Session info

```{r session-info}
sessionInfo()
```
