---
title: "Data Loading: Space Sci-Fi Chapter Codings"
author: "Donald Moratz"
date: "2026-05-25"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
    echo    = TRUE,
    message = FALSE,
    warning = FALSE
)

# Wipe the environment so every knit / run starts from a known state.
rm(list = ls())

# Load shared setup: packages, paths, constants, helper functions.
# source.R lives alongside this .Rmd in scripts/.
source("source.R")

```


# Purpose

This script merges the 46 per-book CSVs produced by the Claude-coding
pipeline into a single tidy dataset suitable for analysis.  The output is
written both wide (one row per chapter, q1..q12 as columns) and long (one
row per chapter x question), in both .rds (preserves types) and .csv
(human-readable) formats, under `data/processed/`.

Decisions encoded here are documented in `PROJECT_TASKS.md` §2.  The major
ones are:

* **Drop paratext rows** matching `paratext_regex` (front matter, intros,
  prefaces, forewords, marked essays, afterwords, appendices).  Prologues
  and epilogues are **kept** as narrative content.
* **Drop one file**: `US_Enders_Game.csv` is superseded by the newer
  `US_Human_Split_Enders_Game.csv`.
* **Override eight book titles** via `book_rename_map` to resolve
  duplicates (Three-Body Problem, Ender's Game, Hyperion, Hitchhiker's
  Omnibus) and normalize cosmetic title-case for four Japanese books.

The entire pipeline is small enough to run in a few seconds on a laptop.

---

# Inputs

## Load the research questions

The questions live in `data/questions.json` and define the canonical answer
options used for validation.  We read them as a list-of-lists (no
auto-simplification) so that nested `answer_options` arrays survive intact.

```{r load-questions}

questions <- jsonlite::read_json(
    QUESTIONS_JSON,
    simplifyVector = FALSE
)

# Convenience vectors for downstream code.
question_numbers <- purrr::map_int(questions, "number")
question_ids     <- paste0("q", question_numbers)

cat("Loaded", length(questions), "questions:\n")
for (q in questions) {
    cat(sprintf("  q%-2d  %s\n", q$number, q$criteria_short))
}

```


## Discover per-book CSV files

We list every `.csv` in `data/results/`, then drop any file in
`file_drop_list` (see §4b of `source.R`).  The remaining files are the
inputs to the rest of the pipeline.

```{r discover-files}

all_csv_paths <- list.files(
    RESULTS_DIR,
    pattern    = "\\.csv$",
    full.names = TRUE
)

kept_csv_paths <- all_csv_paths[
    !basename(all_csv_paths) %in% file_drop_list
]

dropped_files <- setdiff(basename(all_csv_paths), basename(kept_csv_paths))

cat("Total CSV files in data/results/:", length(all_csv_paths), "\n")
cat("Dropped via file_drop_list:     ", length(dropped_files),  "\n")
cat("Files to load:                  ", length(kept_csv_paths), "\n")

if (length(dropped_files) > 0) {
    cat("\nDropped:\n")
    cat(paste0("  ", dropped_files, "\n"), sep = "")
}

```


# Load and combine

## Read every per-book CSV

`load_results_csv()` (defined in `source.R`) reads one CSV and attaches
`source_file` and `file_stem` provenance columns.  All `q*` and
`q*_justification` columns are coerced to character so that readr doesn't
silently guess logical/numeric for low-cardinality columns.

```{r read-all-files}

raw_list <- purrr::map(kept_csv_paths, load_results_csv)

# Sanity-check that every file produced a non-empty tibble.
empty_files <- kept_csv_paths[purrr::map_lgl(raw_list, ~ nrow(.x) == 0)]
if (length(empty_files) > 0) {
    warning(
        "Some CSV files contained no rows: ",
        paste(basename(empty_files), collapse = ", ")
    )
}

cat("Files loaded:        ", length(raw_list), "\n")
cat("Total rows (raw):    ", sum(purrr::map_int(raw_list, nrow)), "\n")

```


## Combine into a single wide tibble

`dplyr::bind_rows()` aligns columns by name.  All per-book CSVs share the
same schema (`country`, `book`, `chapter`, `q1..q12`, `q1_justification..
q12_justification`), so the combined frame has no NA columns from schema
mismatch.

```{r bind-rows}

raw_wide <- dplyr::bind_rows(raw_list)

cat("Combined rows:   ", nrow(raw_wide), "\n")
cat("Combined columns:", ncol(raw_wide), "\n")

# Quick visual sanity check that the schema is what we expect.
cat("\nColumn names:\n")
print(names(raw_wide))

```


## Apply book-title overrides

For each row we look up its `file_stem` in `book_rename_map` (see §4c of
`source.R`).  If the stem is in the map we use the override; otherwise we
keep the `book` column value from the CSV.  `clean_book_name()` is scalar,
so we apply it rowwise (it's run on 45 books worth of rows -- still fast).

```{r normalize-book-names}

renamed_wide <- raw_wide %>%
    dplyr::mutate(
        book = purrr::map2_chr(file_stem, book, clean_book_name)
    )

# Show every (file_stem, final book name) pair so reviewers can spot any
# unintended renames at a glance.
rename_audit <- renamed_wide %>%
    dplyr::distinct(file_stem, book) %>%
    dplyr::arrange(file_stem)

cat("Distinct (file_stem -> book) mappings after rename:\n\n")
print(rename_audit, n = nrow(rename_audit))

```


# Clean

## Drop paratext rows

`drop_paratext()` filters out any row whose `chapter` label starts with
front_matter / Introduction / Preface / Foreword / "Essay -" / Afterword /
Appendix (case-insensitive, leading whitespace ignored).  We record before-
and-after counts per book so we can audit the drop rate in the verification
section below.

```{r drop-paratext}

pre_counts <- renamed_wide %>%
    dplyr::count(country, book, name = "rows_before")

clean_wide <- drop_paratext(renamed_wide)

post_counts <- clean_wide %>%
    dplyr::count(country, book, name = "rows_after")

paratext_summary <- pre_counts %>%
    dplyr::left_join(post_counts, by = c("country", "book")) %>%
    dplyr::mutate(
        rows_after   = tidyr::replace_na(rows_after, 0L),
        rows_dropped = rows_before - rows_after
    ) %>%
    dplyr::arrange(dplyr::desc(rows_dropped), country, book)

cat("Total rows before drop:", nrow(renamed_wide), "\n")
cat("Total rows after drop: ", nrow(clean_wide),   "\n")
cat("Rows dropped:          ", nrow(renamed_wide) - nrow(clean_wide), "\n")

```


## Derive chapter_position and part_position

`parse_chapter_position()` (see §5b of `source.R`) returns a one-row
tibble with `part_position` and `chapter_position`.  We map it over every
chapter label, then `tidyr::unnest_wider()` the two integer columns into
the main frame.

For verification we also keep a list of any labels the parser failed to
recognize -- expected to be empty for the current corpus.

```{r parse-chapter-positions}

clean_wide <- clean_wide %>%
    dplyr::mutate(
        parsed_pos = purrr::map(chapter, parse_chapter_position)
    ) %>%
    tidyr::unnest_wider(parsed_pos)

unparseable_labels <- clean_wide %>%
    dplyr::filter(is.na(chapter_position)) %>%
    dplyr::distinct(country, book, chapter) %>%
    dplyr::arrange(country, book, chapter)

cat("Total chapters parsed:        ", nrow(clean_wide), "\n")
cat(
    "Chapter labels with NA position:",
    nrow(unparseable_labels),
    sprintf(
        "(%.1f%%)\n",
        100 * nrow(unparseable_labels) / nrow(clean_wide)
    )
)

if (nrow(unparseable_labels) > 0) {
    cat("\nDistinct unparseable labels:\n")
    print(unparseable_labels, n = nrow(unparseable_labels))
}

```


# Validate

## Check every answer against the canonical option list

`validate_answers()` walks `questions.json` and flags any cell whose
recorded answer is NOT in that question's `answer_options` list.  An empty
return means every cell is canonical.  This catches typos, off-by-one
edits, and unintended free-text in the coder output.

```{r validate-answers}

answer_issues <- validate_answers(clean_wide, questions)

cat("Answer-validation mismatches:", nrow(answer_issues), "\n")

if (nrow(answer_issues) > 0) {
    cat("\nMismatched rows (showing up to 20):\n")
    print(utils::head(answer_issues, 20))
} else {
    cat("All answers match the canonical option lists. \n")
}

```


# Reshape

## Pivot the clean wide frame to long format

For most analyses (computing per-question response distributions,
intercoder reliability, regressions over question-level outcomes) it is
easier to work with one row per chapter x question.  `pivot_to_long()`
also joins `criteria_short` from `questions.json` so plots can label
panels by the human-readable question text.

```{r pivot-long}

clean_long <- pivot_to_long(clean_wide, questions)

cat("Wide frame:", nrow(clean_wide), "rows x", ncol(clean_wide), "cols\n")
cat("Long frame:", nrow(clean_long), "rows x", ncol(clean_long), "cols\n")
cat(
    "Expected long rows (wide rows x", length(questions), "questions):",
    nrow(clean_wide) * length(questions), "\n"
)

# Glimpse the long frame to confirm column types are sensible.
dplyr::glimpse(clean_long)

```


# Save outputs

## Write .rds and .csv versions of both shapes

We save four files into `data/processed/`:

* `space_chapters_wide.rds` / `.csv` -- one row per chapter, q1..q12
  columns, plus their justifications.
* `space_chapters_long.rds`  / `.csv` -- one row per chapter x question.

`.rds` is the canonical format (preserves integer / character types
exactly).  `.csv` is included for quick inspection in Excel or `head` and
for handoff to non-R collaborators.

```{r save-outputs}

wide_rds <- file.path(PROCESSED_DIR, "space_chapters_wide.rds")
wide_csv <- file.path(PROCESSED_DIR, "space_chapters_wide.csv")
long_rds <- file.path(PROCESSED_DIR, "space_chapters_long.rds")
long_csv <- file.path(PROCESSED_DIR, "space_chapters_long.csv")

saveRDS(clean_wide, wide_rds)
saveRDS(clean_long, long_rds)

readr::write_csv(clean_wide, wide_csv)
readr::write_csv(clean_long, long_csv)

cat("Wrote:\n")
cat(" ", wide_rds, "\n")
cat(" ", wide_csv, "\n")
cat(" ", long_rds, "\n")
cat(" ", long_csv, "\n")

```


# Verification

## Invariants that must hold

These `stopifnot()` calls fail loudly if anything in the pipeline is
malformed.  Run them every knit so regressions are caught immediately.

```{r assertions}

# (1) No NA in the structural identifier columns.
stopifnot(all(!is.na(clean_wide$country)))
stopifnot(all(!is.na(clean_wide$book)))
stopifnot(all(!is.na(clean_wide$chapter)))

# (2) Every q1..q12 column is present.
missing_q_cols <- setdiff(question_ids, names(clean_wide))
stopifnot(length(missing_q_cols) == 0)

# (3) Every q1_justification..q12_justification column is present.
just_ids <- paste0(question_ids, "_justification")
missing_just_cols <- setdiff(just_ids, names(clean_wide))
stopifnot(length(missing_just_cols) == 0)

# (4) Answer validation must be empty (canonical answers only).
stopifnot(nrow(answer_issues) == 0)

# (5) Long frame row count equals wide rows x number of questions.
stopifnot(nrow(clean_long) == nrow(clean_wide) * length(questions))

# (6) The expected 45 books loaded (46 files - 1 dropped via file_drop_list).
expected_books <- length(kept_csv_paths)
loaded_books   <- dplyr::n_distinct(clean_wide$file_stem)
stopifnot(expected_books == loaded_books)

cat("All assertions passed.\n")

```


## Per-book chapter counts (pre and post paratext drop)

A `gt` table showing how many rows each book contributed before and after
the paratext drop.  Books with non-zero `rows_dropped` are highlighted so a
human reviewer can sanity-check that the drop rule fired only on genuine
front-matter / appendix / essay rows.

```{r per-book-counts-table}

# Use max(..., 1) so the color scale's domain is always non-degenerate,
# even on the (unlikely) corpus where nothing is dropped.
max_dropped <- max(paratext_summary$rows_dropped, 1L, na.rm = TRUE)

per_book_counts_table <- paratext_summary %>%
    gt::gt() %>%
    gt::tab_header(
        title    = "Per-book chapter counts before and after paratext drop",
        subtitle = sprintf(
            "%d books loaded, %d rows dropped overall",
            dplyr::n_distinct(paratext_summary$book),
            sum(paratext_summary$rows_dropped)
        )
    ) %>%
    gt::cols_label(
        country      = "Country",
        book         = "Book",
        rows_before  = "Rows (before)",
        rows_after   = "Rows (after)",
        rows_dropped = "Dropped"
    ) %>%
    gt::data_color(
        columns = rows_dropped,
        fn      = scales::col_numeric(
            palette = c("white", "tomato"),
            domain  = c(0, max_dropped)
        )
    )

per_book_counts_table

```


## Distinct labels that were dropped as paratext

A flat list of every chapter label that the drop rule filtered out, with
counts.  Reviewers should scan this to confirm that every line really is
paratext and not a misclassified narrative chapter.

```{r dropped-labels-audit}

dropped_labels <- renamed_wide %>%
    dplyr::filter(stringr::str_detect(chapter, paratext_regex)) %>%
    dplyr::count(country, book, chapter, sort = TRUE)

if (nrow(dropped_labels) == 0) {
    cat("No rows were dropped by paratext_regex.\n")
} else {
    cat(
        "Dropped",
        sum(dropped_labels$n),
        "rows across",
        nrow(dropped_labels),
        "(country, book, label) groups:\n\n"
    )
    print(dropped_labels, n = nrow(dropped_labels))
}

```


## Chapter-position parser coverage

Parser coverage by book.  A book with `pct_parsed == 100` has every
chapter slotted into a sortable integer position.  Books below 100% are
candidates for adding a new strategy to `parse_chapter_position()`.

```{r parser-coverage}

parser_coverage <- clean_wide %>%
    dplyr::group_by(country, book) %>%
    dplyr::summarise(
        n_chapters     = dplyr::n(),
        n_parsed       = sum(!is.na(chapter_position)),
        pct_parsed     = round(100 * n_parsed / n_chapters, 1),
        .groups        = "drop"
    ) %>%
    dplyr::arrange(pct_parsed, country, book)

parser_coverage %>%
    gt::gt() %>%
    gt::tab_header(
        title    = "Chapter-position parser coverage by book",
        subtitle = "Books with <100% have at least one label the parser couldn't slot"
    ) %>%
    gt::cols_label(
        country    = "Country",
        book       = "Book",
        n_chapters = "Chapters",
        n_parsed   = "Parsed",
        pct_parsed = "Parsed (%)"
    )

```


# Human-readable load report

## Write `data/processed/load_report.md`

A standalone markdown summary that a reviewer (or co-author) can read
without launching R.  The report captures everything a human needs to
verify the load succeeded: counts of files, rows, books, drops, parser
coverage, validation status.

```{r write-load-report}

report_path <- file.path(PROCESSED_DIR, "load_report.md")

report_lines <- c(

    "# Data Load Report",
    "",
    sprintf("Generated: %s", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
    sprintf("Script:    scripts/data_loading.Rmd"),
    "",

    "## Files",
    "",
    sprintf("- Total CSVs found:        %d", length(all_csv_paths)),
    sprintf("- Files dropped (skip list): %d", length(dropped_files)),
    sprintf("- Files loaded:            %d", length(kept_csv_paths)),
    "",
    if (length(dropped_files) > 0) {
        c("Dropped files:", "", paste0("- `", dropped_files, "`"), "")
    } else {
        c("No files were skipped via `file_drop_list`.", "")
    },

    "## Rows",
    "",
    sprintf("- Raw rows (after file load):     %d", nrow(renamed_wide)),
    sprintf("- Paratext rows dropped:          %d",
            nrow(renamed_wide) - nrow(clean_wide)),
    sprintf("- Clean wide rows:                %d", nrow(clean_wide)),
    sprintf("- Long-format rows (wide x %d Qs): %d",
            length(questions), nrow(clean_long)),
    "",

    "## Books",
    "",
    sprintf("- Distinct books in clean frame: %d",
            dplyr::n_distinct(clean_wide$book)),
    sprintf("- Distinct countries:            %d",
            dplyr::n_distinct(clean_wide$country)),
    "",

    "Per-country book counts:",
    "",
    paste0(
        "- ",
        clean_wide %>%
            dplyr::distinct(country, book) %>%
            dplyr::count(country) %>%
            dplyr::arrange(country) %>%
            { sprintf("%s: %d books", .$country, .$n) }
    ),
    "",

    "## Validation",
    "",
    sprintf("- Answer-validation mismatches: %d", nrow(answer_issues)),
    sprintf("- Unparseable chapter labels:   %d", nrow(unparseable_labels)),
    "",

    "## Outputs",
    "",
    "Files written to `data/processed/`:",
    "",
    "- `space_chapters_wide.rds`",
    "- `space_chapters_wide.csv`",
    "- `space_chapters_long.rds`",
    "- `space_chapters_long.csv`",
    "- `load_report.md` (this file)",
    ""
)

writeLines(report_lines, report_path)

cat("Wrote ", report_path, "\n", sep = "")

```


# Session info

```{r session-info}

sessionInfo()

```
