---
title: "Intercoder Reliability: Claude vs Human Coding"
author: "Donald Moratz"
date: "2026-05-25"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
    echo    = TRUE,
    message = FALSE,
    warning = FALSE
)

# Wipe the environment so every knit / run starts from a known state.
rm(list = ls())

# Load shared setup (packages, paths, helpers) and IRR-specific helpers.
# Both files live alongside this .Rmd in scripts/.
source("source.R")
source("source_irr.R")

```


# Purpose

This script computes intercoder reliability (IRR) between Claude's per-
chapter coding (from `data/results/`) and the human-coded subset in
`data/human_coded/`.  Five books are double-coded:

* The Martian (long xlsx)
* The Three-Body Problem (long xlsx)
* Planet of the Apes (long xlsx; Q12 not coded by the human)
* Off on a Comet (wide xlsx; first 10 chapters only)
* Ender's Game (long csv; canonicalized 2026-05-25 by Donald)

For every (book, question) pair we report:

* `n_pairs`         -- chapters with non-missing answers from BOTH coders
* `exact_agreement` -- proportion of pairs where Claude == Human
* `kappa`           -- Cohen's unweighted kappa (nominal)
* `kappa (w)`       -- Cohen's squared-weighted kappa (ordinal questions
                       only: Q1-Q8, Q10, Q12)
* `alpha`           -- Krippendorff's alpha (ordinal level for ordinal
                       questions, nominal otherwise)

Thresholds for "acceptable" reliability (from PROJECT_TASKS.md §3):

* kappa  >= `r irr_threshold_kappa`
* alpha  >= `r irr_threshold_alpha`

Both per-(book, question) and pooled-by-question summaries are rendered as
gt tables with pass/fail color coding, and the full pipeline writes a
markdown report to `data/processed/intercoder_reliability_report.md`.

---

# Inputs

## Load the research questions

```{r load-questions}

questions <- jsonlite::read_json(
    QUESTIONS_JSON,
    simplifyVector = FALSE
)

question_numbers <- purrr::map_int(questions, "number")
ordinal_label    <- ifelse(question_numbers %in% ordinal_questions,
                           "ordinal", "nominal")

cat("Loaded", length(questions), "questions",
    "(", sum(ordinal_label == "ordinal"), "ordinal,",
        sum(ordinal_label == "nominal"), "nominal):\n")
for (i in seq_along(questions)) {
    q <- questions[[i]]
    cat(sprintf("  q%-2d  [%s]  %s\n",
                q$number, ordinal_label[i], q$criteria_short))
}

```


## Load Claude's processed wide and long frames

```{r load-claude}

claude_wide <- readRDS(file.path(PROCESSED_DIR, "space_chapters_wide.rds"))
claude_long <- readRDS(file.path(PROCESSED_DIR, "space_chapters_long.rds"))

cat("Claude wide frame:", nrow(claude_wide), "rows x",
    ncol(claude_wide), "cols\n")
cat("Claude long frame:", nrow(claude_long), "rows x",
    ncol(claude_long), "cols\n")
cat("Distinct books in Claude data:", dplyr::n_distinct(claude_wide$book),
    "\n")

```


## Restrict Claude to the 5 IRR books

```{r filter-claude-to-irr}

irr_books <- human_coded_files$book
stopifnot(all(irr_books %in% claude_wide$book))

claude_wide_irr <- claude_wide %>%
    dplyr::filter(book %in% irr_books)

claude_long_irr <- claude_long %>%
    dplyr::filter(book %in% irr_books)

cat("Claude chapters across the 5 IRR books:", nrow(claude_wide_irr), "\n")
cat("Claude (chapter x question) cells:    ", nrow(claude_long_irr), "\n")

claude_wide_irr %>%
    dplyr::count(book, name = "claude_chapters") %>%
    knitr::kable()

```


# Human-coded loaders

## Load and clean all five files

`load_all_human_coded()` runs the schema-specific loader for each book,
applies the approved correction table, coerces NA / blank / non-canonical
values to `"Other / Unsure"`, applies the per-book chapter alignment, and
inner-joins to Claude's per-chapter answers.

```{r load-human}

paired <- load_all_human_coded(claude_wide_irr, claude_long_irr, questions)

cat("Paired (book, chapter, question) rows:", nrow(paired), "\n")

paired %>%
    dplyr::count(book, name = "paired_cells") %>%
    knitr::kable()

```


## Pipeline diagnostics

```{r pipeline-diagnostics}

raw_attr     <- attr(paired, "raw")
cleaned_attr <- attr(paired, "cleaned")
aligned_attr <- attr(paired, "aligned")
residuals    <- attr(paired, "coerced_residuals")

cat("Raw rows (before cleaning):    ", nrow(raw_attr),     "\n")
cat("After NA coercion + corrections:", nrow(cleaned_attr), "\n")
cat("After chapter alignment:       ", nrow(aligned_attr), "\n")
cat("After inner-join to Claude:    ", nrow(paired),       "\n")

if (!is.null(residuals) && nrow(residuals) > 0) {
    cat("\nNon-canonical values coerced to 'Other / Unsure':",
        nrow(residuals), "rows\n")
    residuals %>%
        dplyr::count(book, q_number, mapped_answer,
                     name = "n") %>%
        dplyr::arrange(dplyr::desc(n)) %>%
        head(25) %>%
        knitr::kable(caption = "Top non-canonical values coerced to 'Other / Unsure'")
} else {
    cat("\nNo non-canonical values needed to be coerced.\n")
}

```


## Validate that all human answers are now canonical

```{r validate-human}

remaining <- validate_human_answers(cleaned_attr, questions)

stopifnot(
    "validate_human_answers(): non-canonical mapped_answer remains" =
        nrow(remaining) == 0
)

cat("All cleaned human answers are canonical: OK\n")

```


## Coverage report -- which (book, question) pairs have data?

```{r coverage}

coverage <- paired %>%
    dplyr::count(book, q_number, name = "n_pairs") %>%
    tidyr::pivot_wider(names_from = q_number, values_from = n_pairs,
                       names_prefix = "Q")

knitr::kable(
    coverage,
    caption = paste("Per-(book, question) coverage after cleaning.",
                    "Empty cells indicate the question was not coded for",
                    "that book.")
)

```


# IRR results

## Per-(book, question) metrics

```{r per-book-irr}

irr_per_book <- compute_irr_per_question(paired, questions)

irr_per_book %>%
    format_irr_table(
        caption       = "IRR per (book, question)",
        group_by_book = TRUE
    )

```


## Pooled-by-question metrics

```{r pooled-irr}

irr_overall <- compute_irr_overall(paired, questions)

irr_overall %>%
    format_irr_table(
        caption       = "IRR pooled across all five IRR books, by question",
        group_by_book = FALSE
    )

```


## Corpus-wide pooled summary

Three complementary roll-ups across **all** books and **all** questions:

* **Pooled exact agreement** -- the simple share of cells where Claude
  == Human across all `r nrow(paired)` paired cells.
* **Mean kappa / weighted kappa / alpha** -- arithmetic mean of the
  per-question values from `irr_overall` (each question weighted equally
  regardless of n).
* **Pooled alpha** -- Krippendorff's alpha computed on the whole paired
  frame after re-encoding every `(question_number, answer)` tuple as a
  unique nominal integer.  Treats the whole pipeline as a single nominal
  coding task and gives one chance-corrected number for the corpus.

The one-sided "Other / Unsure" drop still applies to the ordinal
per-question metrics that feed the mean; the pooled-alpha re-encoding
is nominal by construction, so Other / Unsure is just another category.

```{r pooled-overall}

irr_pooled <- compute_irr_pooled(paired, irr_overall)

irr_pooled %>%
    format_irr_pooled_table(
        caption = "Corpus-wide pooled IRR summary"
    )

```


## Threshold pass/fail summary

```{r threshold-summary}

pass_summary <- irr_overall %>%
    dplyr::mutate(
        kappa_pass          = !is.na(kappa) & kappa >= irr_threshold_kappa,
        kappa_weighted_pass = !is.na(kappa_weighted) &
            kappa_weighted >= irr_threshold_kappa,
        alpha_pass          = !is.na(alpha) & alpha >= irr_threshold_alpha
    )

cat("Questions passing kappa >= ", irr_threshold_kappa, ": ",
    sum(pass_summary$kappa_pass), " / ", nrow(pass_summary), "\n", sep = "")
cat("Questions passing weighted kappa >= ", irr_threshold_kappa,
    " (ordinal only): ",
    sum(pass_summary$kappa_weighted_pass), " / ",
    sum(pass_summary$scale == "ordinal"), "\n", sep = "")
cat("Questions passing alpha >= ", irr_threshold_alpha, ": ",
    sum(pass_summary$alpha_pass), " / ", nrow(pass_summary), "\n", sep = "")

failing <- pass_summary %>%
    dplyr::filter(!alpha_pass | !kappa_pass) %>%
    dplyr::select(q_number, n_pairs, exact_agreement,
                  kappa, kappa_weighted, alpha, scale)

if (nrow(failing) > 0) {
    knitr::kable(failing, caption = "Questions BELOW threshold (review).")
} else {
    cat("\nAll questions pass both thresholds.\n")
}

```


# Save outputs

```{r save-outputs}

saveRDS(paired,        file.path(PROCESSED_DIR, "irr_paired.rds"))
saveRDS(irr_per_book,  file.path(PROCESSED_DIR, "irr_per_book.rds"))
saveRDS(irr_overall,   file.path(PROCESSED_DIR, "irr_overall.rds"))
saveRDS(irr_pooled,    file.path(PROCESSED_DIR, "irr_pooled.rds"))

readr::write_csv(paired,
                 file.path(PROCESSED_DIR, "irr_paired.csv"))
readr::write_csv(irr_per_book,
                 file.path(PROCESSED_DIR, "irr_per_book.csv"))
readr::write_csv(irr_overall,
                 file.path(PROCESSED_DIR, "irr_overall.csv"))
readr::write_csv(irr_pooled,
                 file.path(PROCESSED_DIR, "irr_pooled.csv"))

cat("Wrote IRR outputs to", PROCESSED_DIR, "\n")

```


## Generate intercoder_reliability_report.md

A human-readable companion to this knit, written even if the .Rmd is run
non-interactively.

```{r write-report}

report_lines <- c(
    "# Intercoder Reliability Report",
    "",
    paste0("_Generated: ", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "_"),
    "",
    "## Inputs",
    "",
    paste0("- Claude processed long frame: ",
           nrow(claude_long), " (book x chapter x question) cells across ",
           dplyr::n_distinct(claude_wide$book), " books"),
    paste0("- Human-coded books (5): ",
           paste(human_coded_files$book, collapse = "; ")),
    paste0("- Paired (Claude + Human) (book, chapter, question) rows: ",
           nrow(paired)),
    "",
    "## Cleaning summary",
    "",
    paste0("- Raw human-coded long rows: ", nrow(raw_attr)),
    paste0("- After corrections + NA -> 'Other / Unsure': ",
           nrow(cleaned_attr)),
    paste0("- After chapter alignment to Claude: ", nrow(aligned_attr)),
    paste0("- Non-canonical values coerced to 'Other / Unsure': ",
           if (is.null(residuals)) 0 else nrow(residuals)),
    "",
    "## Thresholds",
    "",
    paste0("- kappa >= ", irr_threshold_kappa),
    paste0("- alpha >= ", irr_threshold_alpha),
    "",
    "## Corpus-wide pooled summary",
    "",
    knitr::kable(
        irr_pooled %>%
            dplyr::mutate(dplyr::across(c(pooled_exact_agreement,
                                          mean_kappa, mean_kappa_weighted,
                                          mean_alpha,
                                          pooled_alpha_all,
                                          pooled_alpha_excl_one_other),
                                        ~ round(.x, 3))),
        format = "markdown"
    ),
    "",
    "## Pooled-by-question results",
    "",
    knitr::kable(
        irr_overall %>%
            dplyr::select(-dplyr::any_of(c("n_both_other", "n_one_other",
                                           "n_pairs_ordinal"))) %>%
            dplyr::mutate(dplyr::across(c(exact_agreement, kappa,
                                          kappa_weighted, alpha),
                                        ~ round(.x, 3))),
        format = "markdown"
    )
)

writeLines(
    report_lines,
    file.path(PROCESSED_DIR, "intercoder_reliability_report.md")
)

cat("Wrote intercoder_reliability_report.md to", PROCESSED_DIR, "\n")

```


# Session info

```{r session-info}
sessionInfo()
```
